Exploring Local Explanations of Nonlinear Models Using Animated Linear Projections

### 2022-03-14 _paper.tex
wresting jmlr2e.sty

  pdfkeywords={R, XAI, explanable artificial intelligence, black-box models, local explanations, grand tour, manual tour, radial tour},

@@

\title{Exploring Local Explanations of Nonlinear Models Using Animated Linear Projections}
\author{}
\date{\vspace{-2.5em}}

@@

There are different reasons and emphases to fit a model. Breiman (2001), reiterated by Shmueli (2010), taxonomize modeling based on its purpose; \emph{explanatory} modeling is done for some inferential purpose, while \emph{predictive} modeling focuses more on the predictions of out-of-sample instances. The intended use has important implications for model selection and development. In explanatory modeling, interpretability is vital for drawing inferential conclusions. While predictive modeling may opt for more accurate nonlinear models. The use of black-box models is becoming increasingly common, but not without their share of controversy (O'Neil 2016; Kodiyan 2019). However, the loss of interpretation presents a challenge.

Interpretability is vital for exploring and protecting against potential biases (e.g., sex (Dastin 2018; Duffy 2019), race (Larson et al. 2016), and age (Díaz et al. 2018)) in any model. For instance, models regularly pick up on biases in the training data that have observed influence on the response (output) feature, which is then built into the model. Feature-level (variable-level) interpretability of models is essential in evaluating such biases. It is also generally important for many problems, where it is not enough to predict accurately. Still, one must be able to explain which predictors are most responsible for generating a response value.

Another concern is data drift, a shift in support or domain of the explanatory features (variable or predictors). Nonlinear models are typically more sensitive and do not extrapolate well outside the training data domain. Better interpretability of the model means more transparency where models' predictions may be plausible or completely unreliable.

Explainable Artificial Intelligence (XAI) is an emerging field of research that tries to increase the interpretability of black-box models. A common approach is to use \emph{local explanations}, which attempt to approximate linear feature importance at the location of each instance (observation) or the predictions at a specific point in the data domain. Because these are point-specific, it is challenging to visualize them to comprehensively understand a model.

@@

The remainder of this paper is organized as follows. The following Section, \ref{sec:explanations}, covers the background of the local explanation and the traditional visuals produced. Section \ref{sec:tour} explains the animations of continuous linear projections. Section \ref{sec:cheemviewer} discusses the visual layout in the graphical user interface, how it facilitates analysis, data preprocessing, and package infrastructure. Then Section \ref{sec:casestudies} illustrates the application to supervised learning with categorical and quantitative output. Section \ref{sec:cheemdiscussion} concludes with the insights gained and possible directions to explore in the future.

@@

Following the use case \emph{Explanatory Model Analysis} (Biecek and Burzykowski 2021), FIFA data is used to illustrate SHAP. Consider soccer data from the FIFA 2020 season (Leone 2020). There are 5000 instances of 9 skill measures (after aggregating highly correlated features). A random forest model is fit regressing player's wages {[}2020 Euros{]} from their skill measurements. The SHAP values are compared for a star offensive player (L. Messi) and defensive player (V. van Dijk). The results are displayed in Figure \ref{fig:shapdistrbd}. A difference in the attribution of the feature importance across the two positions of the players can be expected. This would be interpreted as how a player's salary depends on this combination of skill sets. Panel (a) is a modified breakdown plot (Gosiewska and Biecek 2019) where three sequences of features are presented, so the two instances can be more easily compared. The magnitude of the contributions depends on the sequence in which they appear. Panel (b) shows the differences of the player's median values of over 25 such sequences. In summary, these plots highlight how local explanations bring interpretability to a model, at least in the vicinity of their instances. In this instance, two players with different positions receive different profiles of feature importance to explain the prediction of their wages.

@@

To explore the local explanations, coordinated views (Roberts 2007) (also known as ensemble graphics, Unwin and Valero-Mora 2018) are provided in the \emph{cheem viewer} application. There are two primary plots: the \textbf{global view} to give the context of all of the SHAP values and the \textbf{radial tour view} to explore the local explanations with user-controlled rotation. There are numerous user inputs, including feature selection for the radial tour and instance selection for making comparisons. There are different plots used for the categorical and quantitative responses. Figures \ref{fig:classificationcase} and \ref{fig:regressioncase} are screenshots showing the cheem viewer for the two primary tasks: classification (categorical response) and regression (quantitative response).

@@

Visualization is provided by the first two principal components of the data (left) and the attribution (middle) spaces. These single 2D projections will not reveal all of the structure of higher-dimensional space, but they are helpful visual summaries. In addition, a plot of the observed against predicted response values is also provided (Figures \ref{fig:classificationcase}b, \ref{fig:regressioncase}a) to help identify instances poorly predicted by the model. For classification tasks, misclassified instances are circled in red. Linked brushing between the plots is provided, and a tabular display of selected points helps to facilitate exploration of the spaces and the model (shown in Figures \ref{fig:regressioncase}d).

@@

The local explanations for all observations are normalized (squared sum of values adds to 1), and thus, the relative importance of features can be compared across all instances. These are depicted as vertical parallel coordinate plots (Ocagne 1885) on the basis biplot (Gabriel 1971). 1D biplot display depicts the values of the current basis as bars. The parallel coordinate overlays lines connecting one instance's feature attribution (Figures \ref{fig:classificationcase}e and \ref{fig:regressioncase}e). The attribution projections of the PI and CI are shown as dashed and dotted lines. From this plot, the range and density of the importance across all instances can be interpreted. For classification, one would look at differences between groups on any feature. For example, Figure \ref{fig:classificationcase}e suggests that \texttt{bl} is important for distinguishing the green class from the other two. For regression, one might generally observe which features have low values for all instances (not important), for example, \texttt{BMI} and \texttt{pwr} in Figure \ref{fig:regressioncase}e, and which have a range of high and low values (e.g.~\texttt{off}, \texttt{def}) suggesting they are important for some instances and not important for other instances.

@@

The attribution projection of the PI is the initial 1D basis in a radial tour, displayed as a density plot for a categorical response (Figure \ref{fig:classificationcase}f) and as scatterplots for a quantitative response (Figure \ref{fig:regressioncase}f). The PI and CI are indicated by vertical dashed and dotted lines, respectively. The radial tour varies the contribution of the selected feature between 0-1. This is viewed as an animation of the projections from many intermediate bases. Doing so tests the sensitivity of structure (class separation or strength of relationship) to the feature's contribution. For classification, if the separation between classes diminishes when the feature contribution is reduced, this suggests that the feature is important for class separation. For regression, if the relationship scatterplot weakens when the feature contribution is reduced, indicating that the feature is important for accurately predicting the response.

@@

Selecting a misclassified instance as PI and a correctly classified point nearby in data space as CI makes it easier to examine the features most responsible for the error. The global view (Figure \ref{fig:classificationcase}c) displays the model confusion matrix. The radial tour is 1D and displays as density where color indicates class. An animation slider enables the user to vary the contribution of each feature to explore the sensitivity of the separation to that feature.

@@

Selecting an inaccurately predicted instance as PI and an accurately predicted instance with similar feature values as CI is a helpful way to understand how the model is failing or not. The global view (Figure \ref{fig:regressioncase}a) shows a scatterplot of the observed vs predicted values, which should exhibit a strong relationship if the model is a good fit. The points can be colored by a statistic, residual, a measure of outlyingness (log Mahalanobis distance), or correlation aid in understanding the structure identified in these spaces.

@@

The application has several reactive inputs that affect the data used, aesthetic display, and tour manipulation. These reactive inputs make the software flexible and extensible (Figure \ref{fig:classificationcase}a \& d). The application also has more exploratory interactions to help link points across displays, reveal structures found in different spaces, and access the original data.

A tooltip displays instance number/name and classification information while the cursor hovers over a point. Linked brushing allows the selection of points (left click and drag) where those points will be highlighted across plots (Figure \ref{fig:classificationcase}a \& b). The information corresponding to the selected points is populated on a dynamic table (Figure \ref{fig:classificationcase}d). These interactions aid exploration of the spaces and, finally, the identification of primary and comparison instances.

@@

@@

\caption{Examining the SHAP values for a random forest model classifying Palmer penguin species. The PI is a Gentoo (purple) penguin that is misclassified as a Chinstrap (orange), marked as an asterisk in (a), and the dashed vertical line in (b). The radial view shows varying the contribution of \texttt{fl} from the initial attribution projection (b, left), which produces a linear combination where the PI is more probably (higher density value) a Chinstrap than a Gentoo (b, right). (The animation of the radial tour is at \href{https://vimeo.com/666431172}{vimeo.com/666431172}.)}\label{fig:casepenguins}
\end{figure}

Figure \ref{fig:casepenguins} shows plots from the cheem viewer for exploring the random forest model on the penguins data. Panel (a) shows the global view, and panel (b) shows several 1D projections generated with the radial tour. Penguin 243, a Gentoo (purple), is the PI because it has been misclassified as a Chinstrap (orange).

@@

\caption{Checking what is learned from the cheem viewer. This is a plot of flipper length (\texttt{fl}) and bill length (\texttt{bl}), where an asterisk highlights the PI. A Gentoo (purple) misclassified as a Chinstrap (orange). The PI has an unusually small \texttt{fl} length which is why it is confused with a Chinstrap.}\label{fig:casepenguinsblfl}
\end{figure}

@@

Figure \ref{fig:casepenguinsblfl} confirms that flipper length (\texttt{fl}) is vital for the confusion of the PI as a Chinstrap. Here, flipper length and body length are plotted, and the PI can be seen to be closer to the Chinstrap group in these two features, mainly because it has an unusually low value of flipper length relative to other Gentoos. From this view, it makes sense that it is a hard instance to account for as decision trees can only partition only vertical and horizontal lines.

@@

@@

\caption{Examining the local explanation for a PI which is dark (orange) chocolate incorrectly predicted to be milk (green). From the attribution projection, this chocolate correctly looks more like dark than milk, which suggests that the local explanation does not help understand the prediction for this instance. So, the contribution of Sugar is varied --- reducing it corresponds primarily with increasing Fiber. When Sugar is zero, Fiber contributes strongly towards the left. In this particular view, the PI is closer to the bulk of the milk chocolates, suggesting that the prediction put a lot of importance on Fiber. This chocolate is a rare dark chocolate without any Fiber leading to it being mistaken for a milk chocolate. (A video of the tour animation can be found at \href{https://vimeo.com/666431143}{vimeo.com/666431143}.)}\label{fig:casechocolates}
\end{figure}

This type of exploration is shown in Figure \ref{fig:casechocolates}, where a chocolate labeled dark but predicted to be milk is chosen as the PI (instance 22). It is compared with a CI that is a correctly classified dark chocolate (instance 7). The PCA plot and the tree SHAP PCA plots (a) show a big difference between the two chocolate types but with confusion for a handful of instances. The misclassifications are more apparent in the observed vs predicted plot and can be seen to be mistaken in both ways: milk to dark and dark to milk.

@@

It would also be interesting to explore an inverse misclassification. In this case, a milk chocolate is selected while it was misclassified as a dark chocolate. Chocolate 84 is just this case and is compared with a correctly predicted milk chocolate (instance 71). The corresponding global view and radial tour frames are shown in Figure \ref{fig:casechocolatesinverse}.

@@

\caption{Examining the local interpretation for a PI which is milk (green) chocolate incorrectly predicted to be dark (orange). In the attribution projection, the PI could be either milk or dark. Sodium and Fiber have the largest differences in attributed feature importance, with low values relative to other milk chocolates. The lack of importance attributed to these variables is suspected of contributing to the mistake, so the contribution of Sodium is varied. If Sodium had a larger contribution to the prediction (like in this view). the PI would look more like other milk chocolates. (A video of the tour animation can be found at \href{https://vimeo.com/666431148}{vimeo.com/666431148}.)}\label{fig:casechocolatesinverse}
\end{figure}

@@

The 2020 season FIFA data (Leone 2020; Biecek 2018) contains many skill measurements of soccer/football players and wage information. Nine higher-level skill groupings were identified and aggregated from highly correlated features. A random forest model is fit from these predictors, regressing player wages {[}2020 euros{]}. The model was fit from 5000 instances before being thinned to 500 players to mitigate occlusion and render time. Continuing from the exploration in section \ref{sec:explanations}, we are interested to see the difference in attribution based on the exogenous player position. That is, the model should be able to use multiple linear profiles to better predict the wages from different field positions of players despite not having this information. A leading offensive fielder (L. Messi) is compared with a top defensive fielder (V. van Dijk). The same instances were used in figure \ref{fig:shapdistrbd}.

@@

\caption{Exploring the wages {[}euros{]} relative to skill measurements in the FIFA 2020 data. Star offensive player (L. Messi) is the PI and he is compared with a top defensive player (V. van Dijk). The attribution projection is shown at left, and it can be seen that this combination of features produces a view where Messi has very high predicted (and observed) wages. Defense (\texttt{def}) is the chosen feature to vary. It starts off very low, and as its contribution is increased (right plot) it can be seen that Messi's predicted wages decrease dramatically. The increased contribution in defense comes at the expense of offensive and reaction skills. The interpretation is that indeed, Messi's high wages are most attributable to his offensive and reaction skills, as initially provided by the local explanation. (A video of the animated radial tour can be found at \href{https://vimeo.com/666431163}{vimeo.com/666431163}.)}\label{fig:casefifa}
\end{figure}

Figure \ref{fig:casefifa}, tests the support of the local explanation. Offensive and reaction skills (\texttt{off} and \texttt{rct}) are both crucial to explaining a star offensive player. If either of them were rotated out, the other would be rotated into the frame, maintaining a far-right position. However, Increasing the contribution of a variable with low importance would rotate both features out of the frame.

@@

@@

\caption{Exploring an instance with a large residual as the PI from fitting sales price {[}USD{]} to other variables in the Ames housing 2018. The sale price of the PI was under-predicted. The local explanation indicates a sizable attribution to Lot Area (\texttt{LtA}). The CI has a similar predicted sales price and smaller residual and has minimal attribution to Lot Area. In the attribution projection, the PI has a higher sales price than the CI. Reducing the contribution of Lot Area brings these two prices in line. This suggests if the model did not consider Lot Area, then the two houses would be quite similar. That is, the large residual is due to a lack of factoring in the Lot Area for the prediction of PI's sales price. (A video showing the animation is at \href{https://vimeo.com/666431134}{vimeo.com/666431134}.)}\label{fig:caseames}
\end{figure}

Figure \ref{fig:caseames} selects the house sale 74, a sizable under prediction with an enormous Lot Area contribution. The CI has a similar predicted price though the prediction was accurate and gives almost no attribution to lot size. The attribution projection places instances with high Living Areas to the right. The contribution of Living Area contrasts the contribution of this feature. As the contribution of Lot Area decreases, the predictive power decreases for the PI, while the CI remains stationary. This large of importance in the Living Area is relatively uncommon. Boosting tree models may be more resilient to such an under-prediction as they would up-weighting this residual and force its inclusion in the final model.

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\hypertarget{ref-arrieta_explainable_2020}{}%
Arrieta, Alejandro Barredo, Natalia Díaz-Rodríguez, Javier Del Ser, Adrien Bennetot, Siham Tabik, Alberto Barbado, Salvador García, Sergio Gil-López, Daniel Molina, and Richard Benjamins. 2020. {``Explainable {Artificial} {Intelligence} ({XAI}): {Concepts}, Taxonomies, Opportunities and Challenges Toward Responsible {AI}.''} \emph{Information Fusion} 58: 82--115.

@@

\leavevmode\hypertarget{ref-breiman_statistical_2001}{}%
Breiman, Leo. 2001. {``Statistical Modeling: {The} Two Cultures (with Comments and a Rejoinder by the Author).''} \emph{Statistical Science} 16 (3): 199--231.

@@

>>>>

  pdfauthor={Nicholas Spyrison; Dianne Cook; Kimbal Marriott},
  pdfkeywords={R, XAI, explanable artificial intelligence, black-box
models, local explanations, grand tour, manual tour, radial tour},

@@

\title{Exploring Local Explanations of Nonlinear Models Using Animated
Linear Projections}
\author{Nicholas Spyrison \and Dianne Cook \and Kimbal Marriott}
\date{}

@@

There are different reasons and emphases when considering to fit a
model. Breiman (2001b), reiterated by Shmueli (2010), taxonomize models
based on their purpose. \emph{Explanatory} modeling is done for some
inferential purpose, while \emph{predictive} modeling focuses more
narrowly on the performance of some objective function. The intended use
has important implications for model selection and development. In
explanatory modeling, interpretability is vital for drawing inferential
conclusions. Nonlinear models range from additive models with at least
one polynomial term to more complex machine learning models such as
random forest, support-vector machine, or neural network to name a few
(Breiman 2001a; Boser, Guyon, and Vapnik 1992; Anderson 1995,
receptively).

Nonlinear models have many or complex interactive terms, which cause an
opaqueness to the interpretation of the variables. This difficulty in
interpreting the terms in complex nonlinear models sometimes leads to
them being referred to as black box models. Despite the potentially
better performance of nonlinear models, their use is not without
controversy (O'Neil 2016; Kodiyan 2019). And the loss of interpretation
presents a challenge.

Interpretability is vital for exploring and protecting against potential
biases in any model (e.g.~sex -- Dastin (2018); Duffy (2019), race --
Larson et al. (2016), and age -- Díaz et al. (2018)). For instance,
models regularly pick up on biases in the training data where such
classes correlate with changes in the response variable. This bias is
then built into the model. Variable-level (feature-level)
interpretability of models is essential in evaluating and addressing
such biases.

Another concern is data drift, where a shift in the range of the
explanatory variables (features or predictors). Some nonlinear models
are sensitive to this and do not extrapolate well outside the support of
the training data (Espadoto et al. 2021). Maintaining variable
interpretability is also essential to address issues arising from data
drift.

Explainable Artificial Intelligence (XAI) is an emerging field of
research that tries to increase the interpretability of black-box models
(Adadi and Berrada 2018; Arrieta et al. 2020). A common approach is to
use \emph{local explanations}, which attempt to approximate linear
feature importance at the location of each instance (observation) or the
predictions at a specific point in the data domain. Because these are
point-specific, it is challenging to visualize them to comprehensively
understand a model.

@@

The remainder of this paper is organized as follows. The following
Section, @ref(sec:explanations), covers the background of the local
explanation and the traditional visuals produced. Section @ref(sec:tour)
explains the animations of continuous linear projections. Section
@ref(sec:cheemviewer) discusses the visual layout in the graphical user
interface, how it facilitates analysis, data preprocessing, and package
infrastructure. Then Section @ref(sec:casestudies) illustrates the
application to supervised learning with categorical and quantitative
output. Section @ref(sec:cheemdiscussion) concludes with the insights
gained and possible directions to explore in the future.

@@

Following the use case \emph{Explanatory Model Analysis} (Biecek and
Burzykowski 2021), FIFA data is used to illustrate SHAP. Consider soccer
data from the FIFA 2020 season (Leone 2020). There are 5000 instances of
9 skill measures (after aggregating highly correlated features). A
random forest model is fit regressing player's wages {[}2020 Euros{]}
from their skill measurements. The SHAP values are compared for a star
offensive player (L. Messi) and defensive player (V. van Dijk). The
results are displayed in Figure @ref(fig:shapdistrbd). A difference in
the attribution of the feature importance across the two positions of
the players can be expected. This would be interpreted as how a player's
salary depends on this combination of skill sets. Panel (a) is a
modified breakdown plot (Gosiewska and Biecek 2019) where three
sequences of features are presented, so the two instances can be more
easily compared. The magnitude of the contributions depends on the
sequence in which they appear. Panel (b) shows the differences of the
player's median values of over 25 such sequences. In summary, these
plots highlight how local explanations bring interpretability to a
model, at least in the vicinity of their instances. In this instance,
two players with different positions receive different profiles of
feature importance to explain the prediction of their wages.

@@

To explore the local explanations, coordinated views (Roberts 2007)
(also known as ensemble graphics, Unwin and Valero-Mora 2018) are
provided in the \emph{cheem viewer} application. There are two primary
plots: the \textbf{global view} to give the context of all of the SHAP
values and the \textbf{radial tour view} to explore the local
explanations with user-controlled rotation. There are numerous user
inputs, including feature selection for the radial tour and instance
selection for making comparisons. There are different plots used for the
categorical and quantitative responses. Figures
@ref(fig:classificationcase) and @ref(fig:regressioncase) are
screenshots showing the cheem viewer for the two primary tasks:
classification (categorical response) and regression (quantitative
response).

@@

Visualization is provided by the first two principal components of the
data (left) and the attribution (middle) spaces. These single 2D
projections will not reveal all of the structure of higher-dimensional
space, but they are helpful visual summaries. In addition, a plot of the
observed against predicted response values is also provided (Figures
@ref(fig:classificationcase)b, @ref(fig:regressioncase)a) to help
identify instances poorly predicted by the model. For classification
tasks, misclassified instances are circled in red. Linked brushing
between the plots is provided, and a tabular display of selected points
helps to facilitate exploration of the spaces and the model (shown in
Figures @ref(fig:regressioncase)d).

@@

The local explanations for all observations are normalized (squared sum
of values adds to 1), and thus, the relative importance of features can
be compared across all instances. These are depicted as vertical
parallel coordinate plots (Ocagne 1885) on the basis biplot (Gabriel
1971). 1D biplot display depicts the values of the current basis as
bars. The parallel coordinate overlays lines connecting one instance's
feature attribution (Figures @ref(fig:classificationcase)e and
@ref(fig:regressioncase)e). The attribution projections of the PI and CI
are shown as dashed and dotted lines. From this plot, the range and
density of the importance across all instances can be interpreted. For
classification, one would look at differences between groups on any
feature. For example, Figure @ref(fig:classificationcase)e suggests that
\texttt{bl} is important for distinguishing the green class from the
other two. For regression, one might generally observe which features
have low values for all instances (not important), for example,
\texttt{BMI} and \texttt{pwr} in Figure @ref(fig:regressioncase)e, and
which have a range of high and low values (e.g.~\texttt{off},
\texttt{def}) suggesting they are important for some instances and not
important for other instances.

@@

The attribution projection of the PI is the initial 1D basis in a radial
tour, displayed as a density plot for a categorical response (Figure
@ref(fig:classificationcase)f) and as scatterplots for a quantitative
response (Figure @ref(fig:regressioncase)f). The PI and CI are indicated
by vertical dashed and dotted lines, respectively. The radial tour
varies the contribution of the selected feature between 0-1. This is
viewed as an animation of the projections from many intermediate bases.
Doing so tests the sensitivity of structure (class separation or
strength of relationship) to the feature's contribution. For
classification, if the separation between classes diminishes when the
feature contribution is reduced, this suggests that the feature is
important for class separation. For regression, if the relationship
scatterplot weakens when the feature contribution is reduced, indicating
that the feature is important for accurately predicting the response.

@@

Selecting a misclassified instance as PI and a correctly classified
point nearby in data space as CI makes it easier to examine the features
most responsible for the error. The global view (Figure
@ref(fig:classificationcase)c) displays the model confusion matrix. The
radial tour is 1D and displays as density where color indicates class.
An animation slider enables the user to vary the contribution of each
feature to explore the sensitivity of the separation to that feature.

@@

Selecting an inaccurately predicted instance as PI and an accurately
predicted instance with similar feature values as CI is a helpful way to
understand how the model is failing or not. The global view (Figure
@ref(fig:regressioncase)a) shows a scatterplot of the observed vs
predicted values, which should exhibit a strong relationship if the
model is a good fit. The points can be colored by a statistic, residual,
a measure of outlyingness (log Mahalanobis distance), or correlation aid
in understanding the structure identified in these spaces.

@@

The application has several reactive inputs that affect the data used,
aesthetic display, and tour manipulation. These reactive inputs make the
software flexible and extensible (Figure @ref(fig:classificationcase)a
\& d). The application also has more exploratory interactions to help
link points across displays, reveal structures found in different
spaces, and access the original data.

A tooltip displays instance number/name and classification information
while the cursor hovers over a point. Linked brushing allows the
selection of points (left click and drag) where those points will be
highlighted across plots (Figure @ref(fig:classificationcase)a \& b).
The information corresponding to the selected points is populated on a
dynamic table (Figure @ref(fig:classificationcase)d). These interactions
aid exploration of the spaces and, finally, the identification of
primary and comparison instances.

@@

(ref:casepenguins-cap) Examining the SHAP values for a random forest
model classifying Palmer penguin species. The PI is a Gentoo (purple)
penguin that is misclassified as a Chinstrap (orange), marked as an
asterisk in (a), and the dashed vertical line in (b). The radial view
shows varying the contribution of \texttt{fl} from the initial
attribution projection (b, left), which produces a linear combination
where the PI is more probably (higher density value) a Chinstrap than a
Gentoo (b, right). (The animation of the radial tour is at
\href{https://vimeo.com/666431172}{vimeo.com/666431172}.)

@@

\caption{(ref:casepenguins-cap)}\label{fig:casepenguins}
\end{figure}

Figure @ref(fig:casepenguins) shows plots from the cheem viewer for
exploring the random forest model on the penguins data. Panel (a) shows
the global view, and panel (b) shows several 1D projections generated
with the radial tour. Penguin 243, a Gentoo (purple), is the PI because
it has been misclassified as a Chinstrap (orange).

(ref:casepenguinsblfl-cap) Checking what is learned from the cheem
viewer. This is a plot of flipper length (\texttt{fl}) and bill length
(\texttt{bl}), where an asterisk highlights the PI. A Gentoo (purple)
misclassified as a Chinstrap (orange). The PI has an unusually small
\texttt{fl} length which is why it is confused with a Chinstrap.

@@

\caption{(ref:casepenguinsblfl-cap)}\label{fig:casepenguinsblfl}
\end{figure}

@@

Figure @ref(fig:casepenguinsblfl) confirms that flipper length
(\texttt{fl}) is vital for the confusion of the PI as a Chinstrap. Here,
flipper length and body length are plotted, and the PI can be seen to be
closer to the Chinstrap group in these two features, mainly because it
has an unusually low value of flipper length relative to other Gentoos.
From this view, it makes sense that it is a hard instance to account for
as decision trees can only partition only vertical and horizontal lines.

@@

(ref:casechocolates-cap) Examining the local explanation for a PI which
is dark (orange) chocolate incorrectly predicted to be milk (green).
From the attribution projection, this chocolate correctly looks more
like dark than milk, which suggests that the local explanation does not
help understand the prediction for this instance. So, the contribution
of Sugar is varied --- reducing it corresponds primarily with increasing
Fiber. When Sugar is zero, Fiber contributes strongly towards the left.
In this particular view, the PI is closer to the bulk of the milk
chocolates, suggesting that the prediction put a lot of importance on
Fiber. This chocolate is a rare dark chocolate without any Fiber leading
to it being mistaken for a milk chocolate. (A video of the tour
animation can be found at
\href{https://vimeo.com/666431143}{vimeo.com/666431143}.)

@@

\caption{(ref:casechocolates-cap)}\label{fig:casechocolates}
\end{figure}

This type of exploration is shown in Figure @ref(fig:casechocolates),
where a chocolate labeled dark but predicted to be milk is chosen as the
PI (instance 22). It is compared with a CI that is a correctly
classified dark chocolate (instance 7). The PCA plot and the tree SHAP
PCA plots (a) show a big difference between the two chocolate types but
with confusion for a handful of instances. The misclassifications are
more apparent in the observed vs predicted plot and can be seen to be
mistaken in both ways: milk to dark and dark to milk.

@@

It would also be interesting to explore an inverse misclassification. In
this case, a milk chocolate is selected while it was misclassified as a
dark chocolate. Chocolate 84 is just this case and is compared with a
correctly predicted milk chocolate (instance 71). The corresponding
global view and radial tour frames are shown in Figure
@ref(fig:casechocolatesinverse).

(ref:casechocolatesinverse-cap) Examining the local interpretation for a
PI which is milk (green) chocolate incorrectly predicted to be dark
(orange). In the attribution projection, the PI could be either milk or
dark. Sodium and Fiber have the largest differences in attributed
feature importance, with low values relative to other milk chocolates.
The lack of importance attributed to these variables is suspected of
contributing to the mistake, so the contribution of Sodium is varied. If
Sodium had a larger contribution to the prediction (like in this view).
the PI would look more like other milk chocolates. (A video of the tour
animation can be found at
\href{https://vimeo.com/666431148}{vimeo.com/666431148}.)

@@

\caption{(ref:casechocolatesinverse-cap)}\label{fig:casechocolatesinverse}
\end{figure}

@@

The 2020 season FIFA data (Leone 2020; Biecek 2018) contains many skill
measurements of soccer/football players and wage information. Nine
higher-level skill groupings were identified and aggregated from highly
correlated features. A random forest model is fit from these predictors,
regressing player wages {[}2020 euros{]}. The model was fit from 5000
instances before being thinned to 500 players to mitigate occlusion and
render time. Continuing from the exploration in section
@ref(sec:explanations), we are interested to see the difference in
attribution based on the exogenous player position. That is, the model
should be able to use multiple linear profiles to better predict the
wages from different field positions of players despite not having this
information. A leading offensive fielder (L. Messi) is compared with a
top defensive fielder (V. van Dijk). The same instances were used in
figure @ref(fig:shapdistrbd).

(ref:casefifa-cap) Exploring the wages {[}euros{]} relative to skill
measurements in the FIFA 2020 data. Star offensive player (L. Messi) is
the PI and he is compared with a top defensive player (V. van Dijk). The
attribution projection is shown at left, and it can be seen that this
combination of features produces a view where Messi has very high
predicted (and observed) wages. Defense (\texttt{def}) is the chosen
feature to vary. It starts off very low, and as its contribution is
increased (right plot) it can be seen that Messi's predicted wages
decrease dramatically. The increased contribution in defense comes at
the expense of offensive and reaction skills. The interpretation is that
indeed, Messi's high wages are most attributable to his offensive and
reaction skills, as initially provided by the local explanation. (A
video of the animated radial tour can be found at
\href{https://vimeo.com/666431163}{vimeo.com/666431163}.)

@@

\caption{(ref:casefifa-cap)}\label{fig:casefifa}
\end{figure}

Figure @ref(fig:casefifa), tests the support of the local explanation.
Offensive and reaction skills (\texttt{off} and \texttt{rct}) are both
crucial to explaining a star offensive player. If either of them were
rotated out, the other would be rotated into the frame, maintaining a
far-right position. However, Increasing the contribution of a variable
with low importance would rotate both features out of the frame.

@@

(ref:caseames-cap) Exploring an instance with a large residual as the PI
from fitting sales price {[}USD{]} to other variables in the Ames
housing 2018. The sale price of the PI was under-predicted. The local
explanation indicates a sizable attribution to Lot Area (\texttt{LtA}).
The CI has a similar predicted sales price and smaller residual and has
minimal attribution to Lot Area. In the attribution projection, the PI
has a higher sales price than the CI. Reducing the contribution of Lot
Area brings these two prices in line. This suggests if the model did not
consider Lot Area, then the two houses would be quite similar. That is,
the large residual is due to a lack of factoring in the Lot Area for the
prediction of PI's sales price. (A video showing the animation is at
\href{https://vimeo.com/666431134}{vimeo.com/666431134}.)

@@

\caption{(ref:caseames-cap)}\label{fig:caseames}
\end{figure}

Figure @ref(fig:caseames) selects the house sale 74, a sizable under
prediction with an enormous Lot Area contribution. The CI has a similar
predicted price though the prediction was accurate and gives almost no
attribution to lot size. The attribution projection places instances
with high Living Areas to the right. The contribution of Living Area
contrasts the contribution of this feature. As the contribution of Lot
Area decreases, the predictive power decreases for the PI, while the CI
remains stationary. This large of importance in the Living Area is
relatively uncommon. Boosting tree models may be more resilient to such
an under-prediction as they would up-weighting this residual and force
its inclusion in the final model.

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\hypertarget{ref-adadi_peeking_2018}{}%
Adadi, Amina, and Mohammed Berrada. 2018. {``Peeking Inside the
Black-Box: A Survey on Explainable Artificial Intelligence ({XAI}).''}
\emph{IEEE Access} 6: 52138--60.

\leavevmode\hypertarget{ref-anderson_introduction_1995}{}%
Anderson, James A. 1995. \emph{An Introduction to Neural Networks}. MIT
press.

\leavevmode\hypertarget{ref-arrieta_explainable_2020}{}%
Arrieta, Alejandro Barredo, Natalia Díaz-Rodríguez, Javier Del Ser,
Adrien Bennetot, Siham Tabik, Alberto Barbado, Salvador García, Sergio
Gil-López, Daniel Molina, and Richard Benjamins. 2020. {``Explainable
{Artificial} {Intelligence} ({XAI}): {Concepts}, Taxonomies,
Opportunities and Challenges Toward Responsible {AI}.''}
\emph{Information Fusion} 58: 82--115.

@@

\leavevmode\hypertarget{ref-boser_training_1992}{}%
Boser, Bernhard E., Isabelle M. Guyon, and Vladimir N. Vapnik. 1992.
{``A Training Algorithm for Optimal Margin Classifiers.''} In
\emph{Proceedings of the Fifth Annual Workshop on {Computational}
Learning Theory}, 144--52.

\leavevmode\hypertarget{ref-breiman_random_2001}{}%
Breiman, Leo. 2001a. {``Random Forests.''} \emph{Machine Learning} 45
(1): 5--32.

\leavevmode\hypertarget{ref-breiman_statistical_2001}{}%
---------. 2001b. {``Statistical Modeling: {The} Two Cultures (with
Comments and a Rejoinder by the Author).''} \emph{Statistical Science}
16 (3): 199--231.

@@

\leavevmode\hypertarget{ref-espadoto_toward_2021}{}%
Espadoto, Mateus, Rafael M. Martins, Andreas Kerren, Nina S. T. Hirata,
and Alexandru C. Telea. 2021. {``Toward a {Quantitative} {Survey} of
{Dimension} {Reduction} {Techniques}.''} \emph{IEEE Transactions on
Visualization and Computer Graphics} 27 (3): 2153--73.
\url{https://doi.org/10.1109/TVCG.2019.2944182}.
